---
title: "Untitled"
output: html_document
date: "2025-04-09"
---

Install any packages that will be used for this project
```{r, message = FALSE, warning = FALSE}
if(!require(readxl)){install.packages("readxl")}
library(readxl)
if(!require(dplyr)){install.packages("dplyr")}
library(dplyr)
if(!require(lubridate)){install.packages("lubridate")}
library(lubridate)
if(!require(tidyverse)){install.packages("tidyverse")}
library(tidyverse)
```

\text{========================================================================}\
Data Wrangling\
Read the data in to the environment
```{r, message = FALSE, warning = FALSE}
load("data_ml.RData")                   # Load the data
head(data_ml, 6)
```
The first column is a unique identifier for each stock and the second column is the observation date. The other columns are financial or fundamental features for example, Advt_12M_Usd would represents average trading volume over the past 12 months (in USD).\

Cleanning the data
```{r}
data_ml <- data_ml %>%
  distinct() %>% #remove duplicates
  filter(date > "1999-12-31",         # Keep the date with sufficient data points
         date < "2019-01-01") %>%
    arrange(stock_id, date)             # Order the data
head(data_ml,6)
```

We define our train samples as the following:\
Training set: from 2000-01-01 to 2013-12-31\
Validation set: from 2014-01-01 to 2016-12-31\
Testing Set: from 2017-01-01 to 2018-12-31\
```{r}
training_set <- filter(data_ml, date < as.Date("2014-01-15"))
validation_set <- filter(data_ml, (date >= as.Date("2014-01-15") 
                                   & date < as.Date("2017-01-15")))
testing_set <- filter(data_ml, date >= as.Date("2017-01-15"))
```

